---
title: "Autogenous - Selection scan with outFlank"
author: "Luciano V Cosme"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: breezedark
    css:
      - "styles.css"
    toc: yes
    toc_float: no
    toc_depth: 5
editor_options:
  markdown:
    wrap: 120
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval                        = TRUE,
  echo                        = TRUE,
  cache                       = TRUE, # tidy = TRUE,
  class.output                = "bg-success"
)
knitr::opts_knit$set(
  root.dir = rprojroot::find_rstudio_root_file()
)
```


# Overview

In this RMarkdown we will overview how to run local adaptation analysis with R.Sambada

## 1. R libraries and software

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(colorout)
library(here)
library(ggplot2)
# library(Biostrings)
library(scales)
# library(reticulate)
# library(flextable)
# library(officer)
# library(rmdformats)
library(extrafont)
library(forcats)
library(ggrepel)
library(ggtext)
# library(R.SamBada)
library(rgdal)
library(stats)
library(geosphere)
library(vcfR)
library(OutFLANK)
library(ggvenn)
```


## 2. outFLANK with all 3 populations (MAN, NEW and AUT)

This tool does not directly account for population structure. It instead looks for loci that show a higher Fst than the neutral expectation, under a model of hierarchical population structure. The method is robust to hierarchical population structure and a range of demographic histories.

Check the documentation https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html

"OutFLANK is an R package that implements the method developed by Whitlock and Lotterhos (2015) to use likelihood on a trimmed distribution of FST values to infer the distribution of FST for neutral markers. This distribution is then used to assign q-values to each locus to detect outliers that may be due to spatially heterogeneous selection."

Clean env and memory
```{r}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```

Check vignettes
```{r}
browseVignettes("OutFLANK")
```

### 2.1 OutFLANK analysis with quasi-independent set of SNPs

Extract the intergenic SNPs and do LD pruning
```{bash}
plink2 \
--bfile output/quality_control/file7 \
--extract data/files/intergenic_SNPs.txt \
--indep-pairwise 5 1 0.1 \
--out output/outflank/indepSNP \
--silent;
grep 'samples\|variants\|remaining' output/outflank/indepSNP.log
```


Create vcf with intergenic SNPs that are not linked
```{bash}
plink2 \
--bfile output/quality_control/file7 \
--extract output/outflank/indepSNP.prune.in \
--export vcf \
--make-bed \
--maf 0.1 \
--geno 0.2 \
--out output/outflank/intergenic \
--silent;
grep 'samples\|variants\|remaining' output/outflank/intergenic.log
```


We need to import the data

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","intergenic.vcf"))

# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```

We can select the SNPs with low FST to use as neutral SNPs
```{r}
my_fst |>
  filter(FST < 0.2 & FST >=-0.1) |>
  dplyr::select(LocusName) -> neutral_snps

length(neutral_snps$LocusName)

write.table(
  neutral_snps$LocusName,
  file = here("output", "outflank","neutral_SNPs.txt"),
  row.names = FALSE,
  quote = FALSE,
  col.names = FALSE,
  sep = "\n"
)
```


Create a vcf 
```{bash}
plink2 \
--bfile output/outflank/intergenic \
--export vcf \
--extract output/outflank/neutral_SNPs.txt \
--out output/outflank/neutral \
--silent;
grep "samples\|variants" output/outflank/neutral.log 
```

Import the data.

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","neutral.vcf"))
# vcf <- read.vcfR(here("output","outflank","intergenic.vcf"))
# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```


Data prep: decide which SNPs to use for calibrating the null distribution of Fst

```{r}
# Run OutFLANK to estimate the neutral FST distribution
out_trim <- OutFLANK(FstDataFrame = my_fst, Hmin = 0.1, NumberOfSamples = 60, qthreshold = 0.05)
str(out_trim)
```
Check the fit and make sure it looks good, especially in the right tail:
```{r}
OutFLANKResultsPlotter(
  out_trim,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom = FALSE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
## Zoom in on right tail
OutFLANKResultsPlotter(
  out_trim ,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom =
    TRUE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
plot(my_fst$FST, my_fst$FSTNoCorr)
abline(0,1)
```


Also check the P-value histogram:
Here, we plot the “right-tailed” P-values, which means that outliers in the right tail of the FST distribution will have a P-value near zero. Because we ran the algorithm on a trimmed set of SNPs, this will remove some of the signal around selected sites. So we expect this histogram to be flat and maybe have a bump near 0 for selected sites. This histogram looks pretty good.

```{r}
hist(out_trim$results$pvaluesRightTail)
```




### 2.2 Using estimated neutral mean FST and df to calculate P-values for all loci

Now that we’ve estimated neutral mean FST and df to a quasi-independent set of SNPs, we can go and calculate P-values and q-values for all the loci in our dataset before prunning.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

Now we can use the entire data set to estimate the fst

Create a vcf - ld pruned data
```{bash}
plink2 \
--bfile output/quality_control/file7 \
--export vcf \
--extract output/quality_control/indepSNP.prune.in \
--make-bed \
--out output/outflank/autogenous \
--silent;
grep "samples\|variants" output/outflank/autogenous.log 
```

Import the data

```{r}
# Read your vcf file
vcf2 <- read.vcfR(here("output","outflank","autogenous.vcf"))


# Extract genotypes
geno2 <- vcfR::extract.gt(vcf2)

# Extract locus names (positions)
locusNames2 <- vcfR::getPOS(vcf2)

# Samples
sampleNames2 <- colnames(geno2)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames2 <- sapply(strsplit(sampleNames2, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
H <- matrix(NA, nrow = nrow(geno2), ncol = ncol(geno2))

# Recode genotypes
H[geno2 %in% c("0/0", "0|0")] <- 0
H[geno2 %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
H[geno2 %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
H <- replace(H, is.na(H), 9)

# Check the counts of each genotype
table(as.vector(H))

# Extract SNP names
snpNames <- vcfR::getID(vcf2)

# Use the option below for names if you want to make the plot without having to add the postion later on
locusNames2 <- vcfR::getPOS(vcf2)

# Calculate FST
my_fst2 <- MakeDiploidFSTMat(t(H), locusNames = locusNames2, popNames = popNames2)
my_fst3 <- MakeDiploidFSTMat(t(H), locusNames = snpNames, popNames = popNames2) # with SNP id instead of position

# Concatenate the SNP names and the SNP positions
locusNames3 <- paste(snpNames, locusNames2, sep = ".")

# Calculate FST with the combined names
my_fst4 <- MakeDiploidFSTMat(t(H), locusNames = locusNames3, popNames = popNames2)

# Check output
head(my_fst2)
head(my_fst3)
head(my_fst4)
```

Data checks: Heterozygosity vs. FST
Here, you can see how some of the low H loci have high FST. These are all neutral loci in the simulation, and it is important to exclude them from the OutFLANK algorithm.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

```{r}
P1 <-
  pOutlierFinderChiSqNoCorr(
    my_fst4,
    Fstbar = out_trim$FSTNoCorrbar,
    dfInferred = out_trim$dfInferred,
    qthreshold = 0.05,
    Hmin = 0.1
  )
head(P1)
```


```{r}
tail(P1)
```

```{r}
# notice how the output is ordered differently
my_out <- P1$OutlierFlag==TRUE
plot(P1$He, P1$FST, pch=19, col=rgb(0,0,0,0.1))
points(P1$He[my_out], P1$FST[my_out], col="blue")
# check the P-value histogram for the full set of data
# if there are outliers, it should look flat with an inflation near 0
```

```{r}
hist(P1$pvaluesRightTail)
```

Highlight outliers on Manhattan Plot

For publication, we want to show the accurate estimate of FST, not the uncorrected estimate. Remember to exclude those low H loci!

```{r}
# Use the separate() function to split the LocusName column
P1 <- P1 |>
  separate(LocusName, into = c("SNP_id", "LocusName"), sep = "\\.")

# Create plot
plot(P1$LocusName[P1$He>0.1], P1$FST[P1$He>0.1],
     xlab="Position", ylab="FST", col=rgb(0,0,0,0.2))
  points(P1$LocusName[my_out], P1$FST[my_out], col="magenta", pch=20)  
```
output/local_adaptation/file3
We can use our bim file to create a plot with the chromosomal ids
```{r}
#   ____________________________________________________________________________
#   import the bim file with the SNP data                                   ####
snps <-
  read_delim(                    # to learn about the options use here, run ?read_delim on the console.
    here(
      "output", "outflank", "autogenous.bim"
    ),                           # use library here to load it
    col_names      = FALSE,      # we don't have header in the input file
    show_col_types = FALSE,      # suppress message from read_delim
    col_types      = "ccidcc"    # set the class of each column
  )
#
# set column names
colnames(
  snps
) <-                             # to add a header in our tibble
  c(
    "Scaffold", "SNP", "Cm", "Position", "Allele1", "Allele2"
  )
#
# check the tibble
head(snps)
```

Now we can merge it with my_fst3

```{r}
# Use the separate() function to split the LocusName column
my_fst4 <- my_fst4 |>
  separate(LocusName, into = c("SNP_id", "LocusName"), sep = "\\.")

# Merge the data frames
merged_df <- merge(my_fst4, snps, by.x = "SNP_id", by.y = "SNP")

# Rename the column
merged_df <- rename(merged_df, Chromosome = Scaffold)

# View the first few rows of the merged data frame
head(merged_df)
```

Use ggplot to make a new plot
```{r}
# Subset dataframe for He > 0.1
df_sub <- subset(merged_df, He > 0.1)

# Adjust my_out condition to fit merged_df
my_out <- P1$SNP_id[P1$OutlierFlag == TRUE]

# Save it
write.table(
  my_out,
  file = here("output", "outflank","SNPs_outFlank.txt"),
  row.names = FALSE,
  quote = FALSE,
  col.names = FALSE,
  sep = "\n"
)

# Create a new column in the dataframe to define highlight points based on my_out
df_sub$highlight <- ifelse(df_sub$SNP_id %in% my_out, TRUE, FALSE)

# source the plotting function
source(
  here(
    "scripts", "analysis", "my_theme2.R"
  )
)

# Custom label function
k_format <- function(x) {
  paste0(format(x / 1e6, big.mark = "", scientific = FALSE), "k")
}

# Set colors for the chromosomes
color_vector <- c("#CCF6D6", "#F6E1CC", "#CCD8F6")  # Add or remove colors as needed.
names(color_vector) <- unique(df_sub$Chromosome)  # Make sure unique(df_sub$Chromosome) gives all unique Chromosome values.

# Filter data to include only the highest FST value for each chromosome
highest_FST <- df_sub[which(df_sub$highlight == TRUE), ]
highest_FST <- highest_FST[which(highest_FST$FST == ave(highest_FST$FST, highest_FST$Chromosome, FUN = max)), ]

# Create the plot
ggplot(df_sub, aes(x = Position, y = FST)) +
  geom_point(aes(color = Chromosome),
             data = subset(df_sub, highlight == FALSE),
             size = .3) +
  geom_point(
    color = "magenta",
    data = subset(df_sub, highlight == TRUE),
    size = .3
  ) +
  scale_color_manual(values = color_vector, guide = "none") +
  labs(x = "Position", y = "FST", caption = "") +
  facet_wrap(~ Chromosome, scales = "free_x") +
  scale_x_continuous(labels = k_format) +
  my_theme() +
  theme(panel.spacing = unit(.2, "lines"),
        plot.margin = unit(c(1, 1, 2, 2), "lines"),
        plot.caption = element_markdown(face = "italic", color = "#574E4E")) +
  # Annotate only the highest highlighted point for each chromosome
  geom_text_repel(
    data = highest_FST,
    aes(label = SNP_id),
    size = 3,
    nudge_y = 0.01,
    segment.color = NA,
    max.overlaps = Inf # Set to 'Inf' to show all annotations
  )

# Save the plot
ggsave(
  here(
    "output", "outflank", "figures", "outFlank_outliers.pdf"
  ),
  width  = 8,
  height = 5,
  units  = "in"
)
```


## 3. outFLANK with 2 populations (MAN and AUT)

This tool does not directly account for population structure. It instead looks for loci that show a higher Fst than the neutral expectation, under a model of hierarchical population structure. The method is robust to hierarchical population structure and a range of demographic histories.

Check the documentation https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html

"OutFLANK is an R package that implements the method developed by Whitlock and Lotterhos (2015) to use likelihood on a trimmed distribution of FST values to infer the distribution of FST for neutral markers. This distribution is then used to assign q-values to each locus to detect outliers that may be due to spatially heterogeneous selection."

Clean env and memory
```{r}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```


### 3.1 OutFLANK analysis with quasi-independent set of SNPs

Create vcf with intergenic SNPs that are not linked
```{bash}
plink2 \
--bfile output/outflank/man_aut  \
--extract output/outflank/indepSNP.prune.in \
--export vcf \
--make-bed \
--out output/outflank/man_aut2 \
--silent;
grep 'samples\|variants\|remaining' output/outflank/man_aut2.log
```


We need to import the data

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","man_aut2.vcf"))

# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```

We can select the SNPs with low FST to use as neutral SNPs
```{r}
my_fst |>
  filter(FST < 0.2 & FST >=-0.1) |>
  dplyr::select(LocusName) -> neutral_snps

length(neutral_snps$LocusName)

write.table(
  neutral_snps$LocusName,
  file = here("output", "outflank","man_aut2_SNPs.txt"),
  row.names = FALSE,
  quote = FALSE,
  col.names = FALSE,
  sep = "\n"
)
```


Create a vcf 
```{bash}
plink2 \
--bfile output/outflank/man_aut2 \
--export vcf \
--extract output/outflank/man_aut2_SNPs.txt \
--out output/outflank/man_aut3 \
--silent;
grep "samples\|variants" output/outflank/man_aut3.log 
```

Import the data.

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","man_aut3.vcf"))
# vcf <- read.vcfR(here("output","outflank","intergenic.vcf"))
# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```


Data prep: decide which SNPs to use for calibrating the null distribution of Fst

```{r}
# Run OutFLANK to estimate the neutral FST distribution
out_trim <- OutFLANK(FstDataFrame = my_fst, Hmin = 0.1, NumberOfSamples = 60, qthreshold = 0.05)
str(out_trim)
```
Check the fit and make sure it looks good, especially in the right tail:
```{r}
OutFLANKResultsPlotter(
  out_trim,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom = FALSE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
## Zoom in on right tail
OutFLANKResultsPlotter(
  out_trim ,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom =
    TRUE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
plot(my_fst$FST, my_fst$FSTNoCorr)
abline(0,1)
```


Also check the P-value histogram:
Here, we plot the “right-tailed” P-values, which means that outliers in the right tail of the FST distribution will have a P-value near zero. Because we ran the algorithm on a trimmed set of SNPs, this will remove some of the signal around selected sites. So we expect this histogram to be flat and maybe have a bump near 0 for selected sites. This histogram looks pretty good.

```{r}
hist(out_trim$results$pvaluesRightTail)
```




### 3.2 Using estimated neutral mean FST and df to calculate P-values for all loci

Now that we’ve estimated neutral mean FST and df to a quasi-independent set of SNPs, we can go and calculate P-values and q-values for all the loci in our dataset before prunning.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

Now we can use the entire data set to estimate the fst

Create a vcf - ld pruned data
```{bash}
plink2 \
--bfile output/outflank/man_aut \
--export vcf \
--make-bed \
--out output/outflank/man_aut4 \
--silent;
grep "samples\|variants" output/outflank/man_aut4.log 
```

Import the data

```{r}
# Read your vcf file
vcf2 <- read.vcfR(here("output","outflank","man_aut4.vcf"))


# Extract genotypes
geno2 <- vcfR::extract.gt(vcf2)

# Extract locus names (positions)
locusNames2 <- vcfR::getPOS(vcf2)

# Samples
sampleNames2 <- colnames(geno2)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames2 <- sapply(strsplit(sampleNames2, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
H <- matrix(NA, nrow = nrow(geno2), ncol = ncol(geno2))

# Recode genotypes
H[geno2 %in% c("0/0", "0|0")] <- 0
H[geno2 %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
H[geno2 %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
H <- replace(H, is.na(H), 9)

# Check the counts of each genotype
table(as.vector(H))

# Extract SNP names
snpNames <- vcfR::getID(vcf2)

# Use the option below for names if you want to make the plot without having to add the postion later on
locusNames2 <- vcfR::getPOS(vcf2)

# Calculate FST
my_fst2 <- MakeDiploidFSTMat(t(H), locusNames = locusNames2, popNames = popNames2)
my_fst3 <- MakeDiploidFSTMat(t(H), locusNames = snpNames, popNames = popNames2) # with SNP id instead of position

# Concatenate the SNP names and the SNP positions
locusNames3 <- paste(snpNames, locusNames2, sep = ".")

# Calculate FST with the combined names
my_fst4 <- MakeDiploidFSTMat(t(H), locusNames = locusNames3, popNames = popNames2)

# Check output
head(my_fst2)
head(my_fst3)
head(my_fst4)
```

Data checks: Heterozygosity vs. FST
Here, you can see how some of the low H loci have high FST. These are all neutral loci in the simulation, and it is important to exclude them from the OutFLANK algorithm.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

```{r}
P1 <-
  pOutlierFinderChiSqNoCorr(
    my_fst4,
    Fstbar = out_trim$FSTNoCorrbar,
    dfInferred = out_trim$dfInferred,
    qthreshold = 0.05,
    Hmin = 0.1
  )
head(P1)
```


```{r}
tail(P1)
```

```{r}
# notice how the output is ordered differently
my_out <- P1$OutlierFlag==TRUE
plot(P1$He, P1$FST, pch=19, col=rgb(0,0,0,0.1))
points(P1$He[my_out], P1$FST[my_out], col="blue")
# check the P-value histogram for the full set of data
# if there are outliers, it should look flat with an inflation near 0
```

```{r}
hist(P1$pvaluesRightTail)
```

Highlight outliers on Manhattan Plot

For publication, we want to show the accurate estimate of FST, not the uncorrected estimate. Remember to exclude those low H loci!

```{r}
# Use the separate() function to split the LocusName column
P1 <- P1 |>
  separate(LocusName, into = c("SNP_id", "LocusName"), sep = "\\.")

# Create plot
plot(P1$LocusName[P1$He>0.1], P1$FST[P1$He>0.1],
     xlab="Position", ylab="FST", col=rgb(0,0,0,0.2))
  points(P1$LocusName[my_out], P1$FST[my_out], col="magenta", pch=20)  
```
output/local_adaptation/file3
We can use our bim file to create a plot with the chromosomal ids
```{r}
#   ____________________________________________________________________________
#   import the bim file with the SNP data                                   ####
snps <-
  read_delim(                    # to learn about the options use here, run ?read_delim on the console.
    here(
      "output", "outflank", "man_aut.bim"
    ),                           # use library here to load it
    col_names      = FALSE,      # we don't have header in the input file
    show_col_types = FALSE,      # suppress message from read_delim
    col_types      = "ccidcc"    # set the class of each column
  )
#
# set column names
colnames(
  snps
) <-                             # to add a header in our tibble
  c(
    "Scaffold", "SNP", "Cm", "Position", "Allele1", "Allele2"
  )
#
# check the tibble
head(snps)
```

Now we can merge it with my_fst3

```{r}
# Use the separate() function to split the LocusName column
my_fst4 <- my_fst4 |>
  separate(LocusName, into = c("SNP_id", "LocusName"), sep = "\\.")

# Merge the data frames
merged_df <- merge(my_fst4, snps, by.x = "SNP_id", by.y = "SNP")

# Rename the column
merged_df <- rename(merged_df, Chromosome = Scaffold)

# View the first few rows of the merged data frame
head(merged_df)
```

Use ggplot to make a new plot
```{r}
# Subset dataframe for He > 0.1
df_sub <- subset(merged_df, He > 0.1)

# Adjust my_out condition to fit merged_df
my_out <- P1$SNP_id[P1$OutlierFlag == TRUE]

# Save it
write.table(
  my_out,
  file = here("output", "outflank","man_aut_SNPs_outFlank.txt"),
  row.names = FALSE,
  quote = FALSE,
  col.names = FALSE,
  sep = "\n"
)

# Create a new column in the dataframe to define highlight points based on my_out
df_sub$highlight <- ifelse(df_sub$SNP_id %in% my_out, TRUE, FALSE)

# source the plotting function
source(
  here(
    "scripts", "analysis", "my_theme2.R"
  )
)

# Custom label function
k_format <- function(x) {
  paste0(format(x / 1e6, big.mark = "", scientific = FALSE), "k")
}

# Set colors for the chromosomes
color_vector <- c("#CCF6D6", "#F6E1CC", "#CCD8F6")  # Add or remove colors as needed.
names(color_vector) <- unique(df_sub$Chromosome)  # Make sure unique(df_sub$Chromosome) gives all unique Chromosome values.

# Filter data to include only the highest FST value for each chromosome
highest_FST <- df_sub[which(df_sub$highlight == TRUE), ]
highest_FST <- highest_FST[which(highest_FST$FST == ave(highest_FST$FST, highest_FST$Chromosome, FUN = max)), ]

# Create the plot
ggplot(df_sub, aes(x = Position, y = FST)) +
  geom_point(aes(color = Chromosome),
             data = subset(df_sub, highlight == FALSE),
             size = .3) +
  geom_point(
    color = "magenta",
    data = subset(df_sub, highlight == TRUE),
    size = .3
  ) +
  scale_color_manual(values = color_vector, guide = "none") +
  labs(x = "Position", y = "FST", caption = "") +
  facet_wrap(~ Chromosome, scales = "free_x") +
  scale_x_continuous(labels = k_format) +
  my_theme() +
  theme(panel.spacing = unit(.2, "lines"),
        plot.margin = unit(c(1, 1, 2, 2), "lines"),
        plot.caption = element_markdown(face = "italic", color = "#574E4E")) +
  # Annotate only the highest highlighted point for each chromosome
  geom_text_repel(
    data = highest_FST,
    aes(label = SNP_id),
    size = 3,
    nudge_y = 0.01,
    segment.color = NA,
    max.overlaps = Inf # Set to 'Inf' to show all annotations
  )

# Save the plot
ggsave(
  here(
    "output", "outflank", "figures", "man_aut_outFlank_outliers.pdf"
  ),
  width  = 8,
  height = 5,
  units  = "in"
)
```








## 4. outFLANK with 2 populations (NEW and AUT)

This tool does not directly account for population structure. It instead looks for loci that show a higher Fst than the neutral expectation, under a model of hierarchical population structure. The method is robust to hierarchical population structure and a range of demographic histories.

Check the documentation https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html

"OutFLANK is an R package that implements the method developed by Whitlock and Lotterhos (2015) to use likelihood on a trimmed distribution of FST values to infer the distribution of FST for neutral markers. This distribution is then used to assign q-values to each locus to detect outliers that may be due to spatially heterogeneous selection."

Clean env and memory
```{r}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```


### 4.1 OutFLANK analysis with quasi-independent set of SNPs

Create vcf with intergenic SNPs that are not linked
```{bash}
plink2 \
--bfile output/outflank/new_aut  \
--extract output/outflank/indepSNP.prune.in \
--export vcf \
--make-bed \
--out output/outflank/new_aut2 \
--silent;
grep 'samples\|variants\|remaining' output/outflank/new_aut2.log
```


We need to import the data

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","new_aut2.vcf"))

# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```

We can select the SNPs with low FST to use as neutral SNPs
```{r}
my_fst |>
  filter(FST < 0.2 & FST >=-0.1) |>
  dplyr::select(LocusName) -> neutral_snps

length(neutral_snps$LocusName)

write.table(
  neutral_snps$LocusName,
  file = here("output", "outflank","new_aut2_SNPs.txt"),
  row.names = FALSE,
  quote = FALSE,
  col.names = FALSE,
  sep = "\n"
)
```


Create a vcf 
```{bash}
plink2 \
--bfile output/outflank/new_aut2 \
--export vcf \
--extract output/outflank/new_aut2_SNPs.txt \
--out output/outflank/new_aut3 \
--silent;
grep "samples\|variants" output/outflank/new_aut3.log 
```

Import the data.

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","new_aut3.vcf"))
# vcf <- read.vcfR(here("output","outflank","intergenic.vcf"))
# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```


Data prep: decide which SNPs to use for calibrating the null distribution of Fst

```{r}
# Run OutFLANK to estimate the neutral FST distribution
out_trim <- OutFLANK(FstDataFrame = my_fst, Hmin = 0.1, NumberOfSamples = 60, qthreshold = 0.05)
str(out_trim)
```
Check the fit and make sure it looks good, especially in the right tail:
```{r}
OutFLANKResultsPlotter(
  out_trim,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom = FALSE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
## Zoom in on right tail
OutFLANKResultsPlotter(
  out_trim ,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom =
    TRUE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
plot(my_fst$FST, my_fst$FSTNoCorr)
abline(0,1)
```


Also check the P-value histogram:
Here, we plot the “right-tailed” P-values, which means that outliers in the right tail of the FST distribution will have a P-value near zero. Because we ran the algorithm on a trimmed set of SNPs, this will remove some of the signal around selected sites. So we expect this histogram to be flat and maybe have a bump near 0 for selected sites. This histogram looks pretty good.

```{r}
hist(out_trim$results$pvaluesRightTail)
```




### 4.2 Using estimated neutral mean FST and df to calculate P-values for all loci

Now that we’ve estimated neutral mean FST and df to a quasi-independent set of SNPs, we can go and calculate P-values and q-values for all the loci in our dataset before prunning.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

Now we can use the entire data set to estimate the fst

Create a vcf - ld pruned data
```{bash}
plink2 \
--bfile output/outflank/new_aut \
--export vcf \
--make-bed \
--out output/outflank/new_aut4 \
--silent;
grep "samples\|variants" output/outflank/new_aut4.log 
```

Import the data

```{r}
# Read your vcf file
vcf2 <- read.vcfR(here("output","outflank","new_aut4.vcf"))


# Extract genotypes
geno2 <- vcfR::extract.gt(vcf2)

# Extract locus names (positions)
locusNames2 <- vcfR::getPOS(vcf2)

# Samples
sampleNames2 <- colnames(geno2)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames2 <- sapply(strsplit(sampleNames2, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
H <- matrix(NA, nrow = nrow(geno2), ncol = ncol(geno2))

# Recode genotypes
H[geno2 %in% c("0/0", "0|0")] <- 0
H[geno2 %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
H[geno2 %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
H <- replace(H, is.na(H), 9)

# Check the counts of each genotype
table(as.vector(H))

# Extract SNP names
snpNames <- vcfR::getID(vcf2)

# Use the option below for names if you want to make the plot without having to add the postion later on
locusNames2 <- vcfR::getPOS(vcf2)

# Calculate FST
my_fst2 <- MakeDiploidFSTMat(t(H), locusNames = locusNames2, popNames = popNames2)
my_fst3 <- MakeDiploidFSTMat(t(H), locusNames = snpNames, popNames = popNames2) # with SNP id instead of position

# Concatenate the SNP names and the SNP positions
locusNames3 <- paste(snpNames, locusNames2, sep = ".")

# Calculate FST with the combined names
my_fst4 <- MakeDiploidFSTMat(t(H), locusNames = locusNames3, popNames = popNames2)

# Check output
head(my_fst2)
head(my_fst3)
head(my_fst4)
```

Data checks: Heterozygosity vs. FST
Here, you can see how some of the low H loci have high FST. These are all neutral loci in the simulation, and it is important to exclude them from the OutFLANK algorithm.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

```{r}
P1 <-
  pOutlierFinderChiSqNoCorr(
    my_fst4,
    Fstbar = out_trim$FSTNoCorrbar,
    dfInferred = out_trim$dfInferred,
    qthreshold = 0.05,
    Hmin = 0.1
  )
head(P1)
```


```{r}
tail(P1)
```

```{r}
# notice how the output is ordered differently
my_out <- P1$OutlierFlag==TRUE
plot(P1$He, P1$FST, pch=19, col=rgb(0,0,0,0.1))
points(P1$He[my_out], P1$FST[my_out], col="blue")
# check the P-value histogram for the full set of data
# if there are outliers, it should look flat with an inflation near 0
```

```{r}
hist(P1$pvaluesRightTail)
```

Highlight outliers on Manhattan Plot

For publication, we want to show the accurate estimate of FST, not the uncorrected estimate. Remember to exclude those low H loci!

```{r}
# Use the separate() function to split the LocusName column
P1 <- P1 |>
  separate(LocusName, into = c("SNP_id", "LocusName"), sep = "\\.")

# Create plot
plot(P1$LocusName[P1$He>0.1], P1$FST[P1$He>0.1],
     xlab="Position", ylab="FST", col=rgb(0,0,0,0.2))
  points(P1$LocusName[my_out], P1$FST[my_out], col="magenta", pch=20)  
```
output/local_adaptation/file3
We can use our bim file to create a plot with the chromosomal ids
```{r}
#   ____________________________________________________________________________
#   import the bim file with the SNP data                                   ####
snps <-
  read_delim(                    # to learn about the options use here, run ?read_delim on the console.
    here(
      "output", "outflank", "man_aut.bim"
    ),                           # use library here to load it
    col_names      = FALSE,      # we don't have header in the input file
    show_col_types = FALSE,      # suppress message from read_delim
    col_types      = "ccidcc"    # set the class of each column
  )
#
# set column names
colnames(
  snps
) <-                             # to add a header in our tibble
  c(
    "Scaffold", "SNP", "Cm", "Position", "Allele1", "Allele2"
  )
#
# check the tibble
head(snps)
```

Now we can merge it with my_fst3

```{r}
# Use the separate() function to split the LocusName column
my_fst4 <- my_fst4 |>
  separate(LocusName, into = c("SNP_id", "LocusName"), sep = "\\.")

# Merge the data frames
merged_df <- merge(my_fst4, snps, by.x = "SNP_id", by.y = "SNP")

# Rename the column
merged_df <- rename(merged_df, Chromosome = Scaffold)

# View the first few rows of the merged data frame
head(merged_df)
```

Use ggplot to make a new plot
```{r}
# Subset dataframe for He > 0.1
df_sub <- subset(merged_df, He > 0.1)

# Adjust my_out condition to fit merged_df
my_out <- P1$SNP_id[P1$OutlierFlag == TRUE]

# Save it
write.table(
  my_out,
  file = here("output", "outflank","new_aut_SNPs_outFlank.txt"),
  row.names = FALSE,
  quote = FALSE,
  col.names = FALSE,
  sep = "\n"
)

# Create a new column in the dataframe to define highlight points based on my_out
df_sub$highlight <- ifelse(df_sub$SNP_id %in% my_out, TRUE, FALSE)

# source the plotting function
source(
  here(
    "scripts", "analysis", "my_theme2.R"
  )
)

# Custom label function
k_format <- function(x) {
  paste0(format(x / 1e6, big.mark = "", scientific = FALSE), "k")
}

# Set colors for the chromosomes
color_vector <- c("#CCF6D6", "#F6E1CC", "#CCD8F6")  # Add or remove colors as needed.
names(color_vector) <- unique(df_sub$Chromosome)  # Make sure unique(df_sub$Chromosome) gives all unique Chromosome values.

# Filter data to include only the highest FST value for each chromosome
highest_FST <- df_sub[which(df_sub$highlight == TRUE), ]
highest_FST <- highest_FST[which(highest_FST$FST == ave(highest_FST$FST, highest_FST$Chromosome, FUN = max)), ]

# Create the plot
ggplot(df_sub, aes(x = Position, y = FST)) +
  geom_point(aes(color = Chromosome),
             data = subset(df_sub, highlight == FALSE),
             size = .3) +
  geom_point(
    color = "magenta",
    data = subset(df_sub, highlight == TRUE),
    size = .3
  ) +
  scale_color_manual(values = color_vector, guide = "none") +
  labs(x = "Position", y = "FST", caption = "") +
  facet_wrap(~ Chromosome, scales = "free_x") +
  scale_x_continuous(labels = k_format) +
  my_theme() +
  theme(panel.spacing = unit(.2, "lines"),
        plot.margin = unit(c(1, 1, 2, 2), "lines"),
        plot.caption = element_markdown(face = "italic", color = "#574E4E")) +
  # Annotate only the highest highlighted point for each chromosome
  geom_text_repel(
    data = highest_FST,
    aes(label = SNP_id),
    size = 3,
    nudge_y = 0.01,
    segment.color = NA,
    max.overlaps = Inf # Set to 'Inf' to show all annotations
  )

# Save the plot
ggsave(
  here(
    "output", "outflank", "figures", "new_aut_outFlank_outliers.pdf"
  ),
  width  = 8,
  height = 5,
  units  = "in"
)
```








## 5. outFLANK with 2 populations (MAN and NEW)

This tool does not directly account for population structure. It instead looks for loci that show a higher Fst than the neutral expectation, under a model of hierarchical population structure. The method is robust to hierarchical population structure and a range of demographic histories.

Check the documentation https://htmlpreview.github.io/?https://github.com/whitlock/OutFLANK/blob/master/inst/doc/OutFLANKAnalysis.html

"OutFLANK is an R package that implements the method developed by Whitlock and Lotterhos (2015) to use likelihood on a trimmed distribution of FST values to infer the distribution of FST for neutral markers. This distribution is then used to assign q-values to each locus to detect outliers that may be due to spatially heterogeneous selection."

Clean env and memory
```{r}
# Remove all objects from the environment
rm(list = ls())

# Run the garbage collector to free up memory
gc()
```


### 5.1 OutFLANK analysis with quasi-independent set of SNPs

Create vcf with intergenic SNPs that are not linked
```{bash}
plink2 \
--bfile output/outflank/man_new  \
--extract output/outflank/indepSNP.prune.in \
--export vcf \
--make-bed \
--out output/outflank/man_new2 \
--silent;
grep 'samples\|variants\|remaining' output/outflank/man_new2.log
```


We need to import the data

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","man_new2.vcf"))

# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```

We can select the SNPs with low FST to use as neutral SNPs
```{r}
my_fst |>
  filter(FST < 0.2 & FST >=-0.1) |>
  dplyr::select(LocusName) -> neutral_snps

length(neutral_snps$LocusName)

write.table(
  neutral_snps$LocusName,
  file = here("output", "outflank","man_new2_SNPs.txt"),
  row.names = FALSE,
  quote = FALSE,
  col.names = FALSE,
  sep = "\n"
)
```


Create a vcf 
```{bash}
plink2 \
--bfile output/outflank/man_new2 \
--export vcf \
--extract output/outflank/new_aut2_SNPs.txt \
--out output/outflank/man_new3 \
--silent;
grep "samples\|variants" output/outflank/man_new3.log 
```

Import the data.

```{r}
# Read your vcf file
vcf <- read.vcfR(here("output","outflank","man_new3.vcf"))
# vcf <- read.vcfR(here("output","outflank","intergenic.vcf"))
# Extract genotypes
geno <- vcfR::extract.gt(vcf)

# Extract locus names (positions)
locusNames <- vcfR::getPOS(vcf)

# Samples
sampleNames <- colnames(geno)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames <- sapply(strsplit(sampleNames, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

# Recode genotypes
G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
G <- replace(G, is.na(G), 9)

# Check the counts of each genotype
table(as.vector(G))

# Extract SNP names
snpNames <- vcfR::getID(vcf)

# Calculate FST with SNP names
my_fst <- MakeDiploidFSTMat(t(G), locusNames = snpNames, popNames = popNames)

head(my_fst)
```
Check the FST distribution
```{r}
# Create the histogram of FST values
ggplot(my_fst, aes(x=FST)) +
  geom_histogram(binwidth=0.001, fill="blue", alpha=0.7) +
  labs(title="Histogram of FST Values",
       x="FST",
       y="Frequency") +
  theme_minimal()
```


Data prep: decide which SNPs to use for calibrating the null distribution of Fst

```{r}
# Run OutFLANK to estimate the neutral FST distribution
out_trim <- OutFLANK(FstDataFrame = my_fst, Hmin = 0.1, NumberOfSamples = 60, qthreshold = 0.05)
str(out_trim)
```
Check the fit and make sure it looks good, especially in the right tail:
```{r}
OutFLANKResultsPlotter(
  out_trim,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom = FALSE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
## Zoom in on right tail
OutFLANKResultsPlotter(
  out_trim ,
  withOutliers = TRUE,
  NoCorr = TRUE,
  Hmin = 0.1,
  binwidth = 0.001,
  Zoom =
    TRUE,
  RightZoomFraction = 0.05,
  titletext = NULL
)
```


```{r}
plot(my_fst$FST, my_fst$FSTNoCorr)
abline(0,1)
```


Also check the P-value histogram:
Here, we plot the “right-tailed” P-values, which means that outliers in the right tail of the FST distribution will have a P-value near zero. Because we ran the algorithm on a trimmed set of SNPs, this will remove some of the signal around selected sites. So we expect this histogram to be flat and maybe have a bump near 0 for selected sites. This histogram looks pretty good.

```{r}
hist(out_trim$results$pvaluesRightTail)
```




### 5.2 Using estimated neutral mean FST and df to calculate P-values for all loci

Now that we’ve estimated neutral mean FST and df to a quasi-independent set of SNPs, we can go and calculate P-values and q-values for all the loci in our dataset before prunning.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

Now we can use the entire data set to estimate the fst

Create a vcf - ld pruned data
```{bash}
plink2 \
--bfile output/outflank/man_new \
--export vcf \
--make-bed \
--out output/outflank/man_new4 \
--silent;
grep "samples\|variants" output/outflank/man_new4.log 
```

Import the data

```{r}
# Read your vcf file
vcf2 <- read.vcfR(here("output","outflank","man_new4.vcf"))


# Extract genotypes
geno2 <- vcfR::extract.gt(vcf2)

# Extract locus names (positions)
locusNames2 <- vcfR::getPOS(vcf2)

# Samples
sampleNames2 <- colnames(geno2)

# Extract population names by splitting the sample names
# If your sample names contain population information.
popNames2 <- sapply(strsplit(sampleNames2, "_"), "[", 1)

# Initialize new matrix for transformed genotypes
H <- matrix(NA, nrow = nrow(geno2), ncol = ncol(geno2))

# Recode genotypes
H[geno2 %in% c("0/0", "0|0")] <- 0
H[geno2 %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
H[geno2 %in% c("1/1", "1|1")] <- 2

# Replace NA with 9 for missing data
H <- replace(H, is.na(H), 9)

# Check the counts of each genotype
table(as.vector(H))

# Extract SNP names
snpNames <- vcfR::getID(vcf2)

# Use the option below for names if you want to make the plot without having to add the postion later on
locusNames2 <- vcfR::getPOS(vcf2)

# Calculate FST
my_fst2 <- MakeDiploidFSTMat(t(H), locusNames = locusNames2, popNames = popNames2)
my_fst3 <- MakeDiploidFSTMat(t(H), locusNames = snpNames, popNames = popNames2) # with SNP id instead of position

# Concatenate the SNP names and the SNP positions
locusNames3 <- paste(snpNames, locusNames2, sep = ".")

# Calculate FST with the combined names
my_fst4 <- MakeDiploidFSTMat(t(H), locusNames = locusNames3, popNames = popNames2)

# Check output
head(my_fst2)
head(my_fst3)
head(my_fst4)
```

Data checks: Heterozygosity vs. FST
Here, you can see how some of the low H loci have high FST. These are all neutral loci in the simulation, and it is important to exclude them from the OutFLANK algorithm.

Note that it is important to run this code with the uncorrected FSTs (FSTNoCorr) and the uncorrected mean FST (FSTNoCorrbar).

```{r}
P1 <-
  pOutlierFinderChiSqNoCorr(
    my_fst4,
    Fstbar = out_trim$FSTNoCorrbar,
    dfInferred = out_trim$dfInferred,
    qthreshold = 0.05,
    Hmin = 0.1
  )
head(P1)
```


```{r}
tail(P1)
```

```{r}
# notice how the output is ordered differently
my_out <- P1$OutlierFlag==TRUE
plot(P1$He, P1$FST, pch=19, col=rgb(0,0,0,0.1))
points(P1$He[my_out], P1$FST[my_out], col="blue")
# check the P-value histogram for the full set of data
# if there are outliers, it should look flat with an inflation near 0
```

```{r}
hist(P1$pvaluesRightTail)
```

Highlight outliers on Manhattan Plot

For publication, we want to show the accurate estimate of FST, not the uncorrected estimate. Remember to exclude those low H loci!

```{r}
# Use the separate() function to split the LocusName column
P1 <- P1 |>
  separate(LocusName, into = c("SNP_id", "LocusName"), sep = "\\.")

# Create plot
plot(P1$LocusName[P1$He>0.1], P1$FST[P1$He>0.1],
     xlab="Position", ylab="FST", col=rgb(0,0,0,0.2))
  points(P1$LocusName[my_out], P1$FST[my_out], col="magenta", pch=20)  
```

No outliers for MAN and NEW

## 6. Venn diagrams


```{r, fig.height=5, fig.width=5}
# Read data from txt files
MAN_AUT <-
  read.table(
    here("output", "outflank", "man_aut_SNPs_outFlank.txt"),
    stringsAsFactors = FALSE
  ) |>
  drop_na()


NEW_AUT <-
  read.table(
    here("output", "outflank", "new_aut_SNPs_outFlank.txt"),
    stringsAsFactors = FALSE
  )

NEW_MAN_AUT <-
  read.table(
    here("output", "outflank", "SNPs_outFlank.txt"),
    stringsAsFactors = FALSE
    )


# Create a list with all dataframes
list_of_clusters <- list("MAN_AUT" = MAN_AUT$V1, "NEW_AUT" = NEW_AUT$V1, "NEW_MAN_AUT" = NEW_MAN_AUT$V1)


# Create Venn diagram
venn_diagram <- ggvenn(list_of_clusters, fill_color = c("steelblue", "darkorange", "pink"))
print(venn_diagram)

# Find common SNPs
common_SNPs <- Reduce(intersect, list_of_clusters)

# # Save the shared SNP ids into a txt file
write.table(
  common_SNPs,
  file = here("output", "pcadapt", "common_SNPs_NEW_MAN_AUT.txt"),
  row.names = FALSE,
  col.names = FALSE,
  quote = FALSE
)

# # # Save Venn diagram to PDF
output_path <- here("output", "pcadapt", "figures", "significant_snps_NEW_MAN_AUT.pdf")
ggsave(output_path, venn_diagram, height = 5, width = 5, dpi = 300)
```




